Efficient Thresholded Correlation using Truncated Singular Value Decomposition

### 2015-07-22 threshold/draft.Rnw
Added notes on general matrices (scaling and centering)

\begin{lemma}
Let $a_i$ and $a_j$, $1\le i,j\le n$ be any two columns of the matrix
$A$. Then
\begin{equation}\label{cordist}
\mathrm{cor}(a_i,a_j) = a_i^Ta_j = 1 - \|a_i - a_j\|^2/2.
\end{equation}
In particular, for a given correlation threshold $t$, $0<t<1$, pairs
of vectors $a_i$ and $a_j$ such that $\|a_i - a_j\|^2 > 2(1-t)$ lie
below the correlation threshold.
\end{lemma}
\begin{proof} (XXX perhaps omit this)
\begin{eqnarray*}
\|a_i - a_j\|^2 &=& (a_i - a_j)^T(a_i - a_j) \\
&=& a_i^T a_i + a_j^T a_j - 2a_i^T a_j \\
&=& \|a_i\|^2 + \|a_j\|^2 - 2a_i^T a_j \\
&=& 2 - 2a_i^T a_j.
\end{eqnarray*}
The lemma follows by solving for $a_i^Ta_j$.
\end{proof}

@@

We have so far assumed that the mean of each column vector of $A$ is zero and
the columns are scaled to have unit norm, but we really want a method that works
with general matrices $A$. We've also implicitly assumed that we can
efficiently compute a relatively low-rank truncated SVD of potentially
large  matrices required by step 1 of Algorithm \ref{prune}.  Fortunately,
both needs are met by one algorithm.

The augmented implicitly restarted Lanczos bidiagonalization algorithm (IRLBA)
of Baglama and Reichel \cite{br} efficiently computes truncated singular value
decompositions of large dense or sparse matrices. Reference implementations are
available for R\cite{irlba-R}, Matlab\cite{irlba-M}, and Python\cite{irlba-P}.
The R implementation includes efficient methods for re-starting the algorithm
to increase the subspace dimension (see step 5 of Algorithm \ref{prune}), and
efficient methods for computing the SVD of scaled and centered matrices without
explicitly forming a new matrix.

Let $x\in\mathcal{R}^n$ represent the vector of column means of the matrix $A$,
$W\in\mathcal{R}^{n\times n}$ be a diagonal matrix with the reciprocal of the
column norms of $A$ along the diagonal, and let $e\in\mathcal{R}^m$ be a vector
of all ones. Then $(A - ex^T)W$ is a centered matrix with zero column means,
and scaled to have unit column norms, and

XXX

IRLBA, based on the Lanczos process, is an iterative method based on
matrix-vector products involving the matrix $A$.

XXX scaling/centering discussion and example

>>>>

\begin{lemma}
Let $a_i$ and $a_j$, $1\le i,j\le n$ be any two columns of the matrix
$A$. Then
\begin{equation}\label{cordist}
\mathrm{cor}(a_i,a_j) = a_i^Ta_j = 1 - \|a_i - a_j\|^2/2.
\end{equation}
In particular, for a given correlation threshold $t$, $0<t<1$, pairs
of vectors $a_i$ and $a_j$ such that $\|a_i - a_j\|^2 > 2(1-t)$ lie
below the correlation threshold.
\end{lemma}
\begin{proof} (XXX probably omit this)
\begin{eqnarray*}
\|a_i - a_j\|^2 &=& (a_i - a_j)^T(a_i - a_j) \\
&=& a_i^T a_i + a_j^T a_j - 2a_i^T a_j \\
&=& \|a_i\|^2 + \|a_j\|^2 - 2a_i^T a_j \\
&=& 2 - 2a_i^T a_j.
\end{eqnarray*}
The lemma follows by solving for $a_i^Ta_j$.
\end{proof}

@@

We have so far assumed that the mean of each column vector of the $m\times n$
matrix $A$ is zero and the columns are scaled to have unit norm, but we really
want a method that works with general matrices $A$. We've also implicitly
assumed that we can efficiently compute a relatively low-rank truncated SVD of
potentially large  matrices required by step 1 of Algorithm \ref{prune}.
Fortunately, both needs are met by one algorithm.

The augmented implicitly restarted Lanczos bidiagonalization algorithm (IRLBA)
of Baglama and Reichel \cite{br} efficiently computes truncated singular value
decompositions of large dense or sparse matrices. Reference implementations are
available for R\cite{irlba-R}, Matlab\cite{irlba-M}, and Python\cite{irlba-P}.
The R implementation includes easy to use and efficient methods for restarting
to increase the subspace dimension (see step 5 of Algorithm \ref{prune}), and
for computing the SVD of scaled and centered matrices without explicitly
forming a new matrix.

We relax the column mean and scale assumptions without much additional
computational or storage overhead as follows.  Let
$x=[x_1,x_2,\ldots,x_n]\in\mathcal{R}^n$ represent the vector of column means
of the matrix $A$, $w=[w_1^2,w_2^2,\ldots,w_n^2]\in\mathcal{R}^n$ be the vector
of squared column norms of $A$, and $W\in\mathcal{R}^{n\times n}$ a diagonal
matrix with diagonal entries $1/\sqrt{w_i^2 - m x_i^2}$.  Let
$e\in\mathcal{R}^m$ be a vector of all ones.  Then $(A - ex^T)W$ is a centered
matrix with zero column means, and scaled to have unit column norms, and
\[
\mathrm{cor}(A) = W^T (A-ex^T)^T (A-ex^T) W.
\]
IRLBA, based on the Lanczos process, is a method of iterated matrix-vector
products. The main idea behind efficient application of IRLBA to correlation
problems replaces matrix-vector products of the form $Az$ with $AWz - ex^TWz$ in
the iterations, implicitly working with a scaled and centered matrix without
forming it.  This comes at the cost of storing two additional length $n$ vectors and
computing two additional vector inner products per matrix-vector product.
